Bayesian model comparison for rare variant association studies of multiple phenotypes

### 2018-04-14 tex/appendix.tex

%\section*{Appendix S1}
{\Huge
\textbf\newline{Appendix S1} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline

@@

In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$. Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 1$) than missense variants ($\sigma = 0.1$). 
Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$.

>>>>

%\section*{Appendix S1}
{\Huge
\textbf\newline{Appendix S1} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}

{\LARGE
\textbf\newline{Manuel A. Rivas and Matti Pirinen} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
{\large
\textbf\newline{January 28, 2018} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}

%\newline

@@

In a single study, the prior density for ${\bs {\beta}}$ incorporates the expected correlation of genetic effects among 
a group of variants ($\mathbf{R}_{\textrm{var}}$) and among a group of phenotypes ($\mathbf{R}_{\textrm{phen}}$). 
In addition, we  incorporate an expected spread of the effect size of each variant by scaling $\mathbf{R}_{\textrm{var}}$ as 
$$\mathbf{S}_{\textrm{var}} = \Delta\left(\sigma_m\right) \mathbf{R}_{\textrm{var}} \Delta\left(\sigma_m\right),$$
where $\Delta\left(\sigma_m\right)$ is a diagonal matrix with entries $\sigma_m$ determining the spread of the effect size distribution for each variant $m \leq M$. Thus, we can model settings where, e.g., protein-truncating variants have larger effect sizes ($\sigma = 0.5$) than missense variants ($\sigma = 0.2$). 
Note that when $\sigma_m = 1$ for all $m$ then $\mathbf{S}_{\textrm{var}} = \mathbf{R}_{\textrm{var}}$.

### 2018-04-14 tex/manuscript.tex

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. We consider the use of summary statistic data to apply univariate and multivariate gene-based meta-analysis models for identifying rare variant associations with an emphasis on protective protein-truncating variants that can expedite drug discovery. Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals. We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis (43,626 cases), eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis (5,863 cases), intra-ocular pressure, and corneal resistance factor. We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.3$). We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$). Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.

@@

\subsubsection*{Asthma, eosinophil counts, forced expiratory volume, and forced vital capacity}
We first applied MRP to GWAS summary statistics for asthma, eosinophil count, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC) phenotypes. Recent work has identified associations between the PTV rs146597587 in \textit{IL33} and asthma and eosinophil counts \cite{DeBoever179762, 10.1371/journal.pgen.1006659}. FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}. To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to rare missense variants and PTVs (MAF $< 1\%$) for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene. We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}. In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis (\nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}. We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association (Fig~\ref{asthma_bma}F). The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}). We inspected the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and found that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).

@@

\subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses. Intra-ocular pressure is a measure of the fluid pressure in the eye, is associated with glaucoma risk, and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}. Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}. While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} and \textit{WNT10A} as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}). Applying the MRP model identified \textit{ANGPTL7} with effects consistent with protection to glaucoma, and additional association to \textit{WNT10A} (Fig~\ref{glaucoma_bma}E). We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).

Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}. The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that the association with glaucoma is driven by the missense variant rs28991009 that changes residue 175 from glutamine to histidine (Fig~\ref{glaucoma_bma}F, G). According to the HIPED protein expression database, \textit{ANGPTL7} protein is expressed at $\sim0.7$ parts per million in vitreous humor, the material between the lens and retina of the eyeball; in contrast, the expression of \textit{ANGPTL7} protein is less than $0.01$ parts per million in 68 other normal tissues \cite{Fishilevich:2016hx}. Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target. \textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases.

@@

% %% TABLE: GLAUCOMA GENES OF INTEREST
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
%\caption{
%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
\hline
{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
\thickhline
\textit{ANGPTL7} &              11.9 &  11.9 &      1.7 &                                       10.1 &                       6.2 \\
\hline
\textit{KLHL22}  &               3.5 &   3.5 &     -0.2 &                                        3.9 &                       2.2 \\
\hline
\textit{WNT10A}  &               2.5 &   2.5 &     -0.3 &                                       -0.5 &                       3.5 \\
\hline
\textit{GLI3}    &               1.5 &   1.5 &     -0.3 &                                       -0.7 &                       3.2 \\
\hline
\textit{RNASE2}  &              -0.5 &  -0.5 &      0.4 &                                       -1.0 &                      -1.1 \\
\hline
\textit{FUK}     &              -1.8 &  -1.8 &      2.5 &                                       -2.1 &                      -2.1 \\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}
\paragraph*{S2 Table.}
\label{glaucoma_table}
{\bf Highlighted genes from glaucoma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Whole genome sequencing studies applied to large populations or biobanks with extensive phenotyping raise new analytic challenges. The need to consider many variants at a locus or group of genes simultaneously and the potential to study many correlated phenotypes with shared genetic architecture provide opportunities for discovery and inference that are not addressed by the traditional one variant-one phenotype association study. Here we introduce a model comparison approach we refer to as MRP for rare variant association studies that considers correlation, scale, and location of genetic effects across a group of genetic variants, phenotypes, and studies. We consider the use of summary statistic data to apply univariate and multivariate gene-based meta-analysis models for identifying rare variant associations with an emphasis on protective protein-truncating variants that can expedite drug discovery. Through simulation studies, we demonstrate that the proposed model comparison approach can improve ability to detect rare variant association signals. We also apply the model to two groups of phenotypes from the UK Biobank: 1) asthma diagnosis (43,626 cases), eosinophil counts, forced expiratory volume, and forced vital capacity; and 2) glaucoma diagnosis (5,863 cases), intra-ocular pressure, and corneal resistance factor. We are able to recover known associations such as the protective association between rs146597587 in \textit{IL33} and asthma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 29.4$). We also find evidence for novel protective associations between rare variants in \textit{ANGPTL7} and glaucoma ($\log_{10} \left( \textrm{Bayes Factor}\right) = 11.9$). Overall, we show that the MRP model comparison approach is able to retain and improve upon useful features from widely-used meta-analysis approaches for rare variant association analyses and prioritize protective modifiers of disease risk.

@@

\subsubsection*{Asthma, eosinophil counts, forced expiratory volume, and forced vital capacity}
We first applied MRP to GWAS summary statistics for asthma, eosinophil count, forced expiratory volume in 1-second (FEV$_1$), and forced vital capacity (FVC) phenotypes. Recent work has identified associations between the PTV rs146597587 in \textit{IL33} and asthma and eosinophil counts \cite{DeBoever179762, 10.1371/journal.pgen.1006659}. FEV$_1$ and FVC are measures of pulmonary function that are used to diagnosis and classify pulmonary disease \cite{Swanney1046}. To demonstrate the advantage of considering the phenotypes jointly, we applied MRP to rare missense variants and PTVs (MAF $< 1\%$) for each phenotype separately (Fig~\ref{asthma_bma}A-D) as well as to all phenotypes jointly (Fig~\ref{asthma_bma}E,F) and obtained $\log_{10}$ BF for each gene. We applied both independent and similar effects models and used Bayesian model averaging to compute a single BF per gene \cite{10.2307/2676803}. In agreement with previous studies, we observed evidence that rare missense variants and/or PTVs in \textit{IL33} affect eosinophil counts and offer protection from asthma from the single-phenotype analyses, though the evidence of association was strongest for the joint analysis ($\log_{10} \textrm{BF} = 29.3$, \nameref{asthma_table}) \cite{DeBoever179762, 10.1371/journal.pgen.1006659}. We performed an analysis focused on identifying protective variants which also identified the \textit{IL33} association ($\log_{10} \textrm{BF} = 29.4$, Fig~\ref{asthma_bma}F). The results were similar using only either the independent effects (\nameref{asthma_independent}) or similar effects models (\nameref{asthma_similar}). We inspected the effect sizes from the marginal GWAS regressions for the rare variants included in the analysis and found that the association identified by MRP is likely driven by the PTV rs146597587 (Fig~\ref{asthma_bma}G).

@@

\subsubsection*{Glaucoma, intra-ocular pressure, and corneal resistance factor}
We also applied MRP to missense variants and PTVs for glaucoma, intra-ocular pressure, and corneal resistance factor as well as performing joint analyses. Intra-ocular pressure is a measure of the fluid pressure in the eye, is associated with glaucoma risk, and has been linked to genetic variants associated with glaucoma \cite{doi:10.1001/jama.2014.3192}. Corneal resistance factor is a measure of the cornea's ability to resist mechanical stress and has been associated with glaucoma presence and severity \cite{CXO:CXO414, MANSOURI2012419, GriseDulac:2012bo}. While the individual glaucoma analysis did not yield any associations with $\log_{10}$ BF greater than three, the joint analysis identified rare coding variants in both \textit{ANGPTL7} ($\log_{10} \textrm{BF} = 11.9$) and \textit{WNT10A} ($\log_{10} \textrm{BF} = 2.5$) as associated with glaucoma (Fig~\ref{glaucoma_bma}A-D, \nameref{glaucoma_table}). Applying the protective MRP model also identified the protective association for \textit{ANGPTL7} against glaucoma and added support for an association for \textit{WNT10A} (Fig~\ref{glaucoma_bma}E). We obtained similar results using the independent effects (\nameref{glaucoma_independent}) or similar effects models (\nameref{glaucoma_similar}).

Expression of \textit{ANGPTL7} is upregulated in glaucoma and has been proposed to regulate intra-ocular pressure and glaucoma risk \cite{Comes:2011ex, Kuchtey:2008ek}. The GWAS summary statistics for the rare variants in \textit{ANGPTL7} suggest that the association with glaucoma is driven by the missense variant rs28991009 that changes residue 175 from glutamine to histidine (Fig~\ref{glaucoma_bma}F, G). According to the HIPED protein expression database, \textit{ANGPTL7} protein is expressed at $\sim0.7$ parts per million in vitreous humor, the material between the lens and retina of the eyeball; in contrast, the expression of \textit{ANGPTL7} protein is less than $0.01$ parts per million in 68 other normal tissues \cite{Fishilevich:2016hx}. Such tissue-specific activity may make \textit{ANGPTL7} a useful therapeutic target. \textit{WNT10A} has not been previously associated with glaucoma though an exonic variant rs121908120 in \textit{WNT10A} is associated with central cornea thickness and increased risk of keratoconus, a disease of the cornea, indicating that this gene may play a role in ocular diseases \cite{CuellarPartida:2015hb}.

@@

% %% TABLE: GLAUCOMA GENES OF INTEREST
\begin{table}[!ht]
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
%\caption{
%{\bf $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.}}
\begin{tabular}{|l+p{3cm}|p{1cm}|p{1.6cm}|p{3.8cm}|p{3.4cm}|}
\hline
{\bf Gene} & {\bf Joint, protective} & {\bf Joint} & {\bf Glaucoma} & {\bf Intra-ocular pressure, Goldmann-correlated} & {\bf Corneal resistance factor} \\
\thickhline
\textit{PGLYRP4} &              23.9 &   7.6 &     -0.1 &                                       -1.0 &                      -1.1 \\
\hline
\textit{ANGPTL7} &              12.9 &  12.2 &      1.7 &                                       10.5 &                       6.4 \\
\hline
\textit{JPH4}    &               6.4 &   2.7 &      0.6 &                                       -1.1 &                      -1.2 \\
\hline
\textit{WNT10A}  &               3.5 &   4.1 &     -0.2 &                                       -0.6 &                       3.1 \\
\hline
\textit{KLHL22}  &               3.1 &   3.7 &     -0.2 &                                        3.9 &                       2.2 \\
\hline
\textit{GLI3}    &               1.4 &   2.5 &     -0.4 &                                       -0.1 &                       4.3 \\
\hline
\textit{C4orf29} &              -0.0 &   6.4 &     -0.1 &                                       -0.7 &                      -1.1 \\
\hline
\end{tabular}
\end{adjustwidth}
\end{table}
\paragraph*{S2 Table.}
\label{glaucoma_table}
{\bf Highlighted genes from glaucoma analysis.} $\log_{10}$ Bayes Factors for genes highlighted in Figure~\ref{glaucoma_bma}.
